Dynamical Properties of Molecular Cloud Complexes in Galaxies at the Epoch of Reionization

### 2019-03-05 WRITEUP/ms.tex
minor text edit

\begin{abstract}
We study the dynamical properties of molecular clouds complexes (MC) in a prototypical galaxy (\flower)
and its satellite galaxies at the Epoch of Reionization (EoR) using cosmological zoom-in simulation.
We identify MCs using H$_2$ density, 
and compare their masses, sizes, velocity dispersions, gas surface densities, and virial parameters ($\alpha_{\rm vir}$)
with those observed in nearby and \z$\sim$\,2 galaxies.
The MCs span a mass range of $10^{6.5-9}$\,\Msun. Their velocity dispersions and gas surface densities are 
systematically higher than those observed in nearby galaxies, but comparable to starburst galaxies.
We compute the effective two-component Toomre $Q$ parameter (gas and stars) to examine 
regions that are unstable to local axisymmetric perturbations. 
We further perform virial analysis, which is accessible from observations, to examine regions likely to be undergo collapse.
Our analysis indicates that the MCs in the main disk of \flower are unbound, but its 
substructures have lower virial parameters, which is
consistent with the notion that collapsing structures result from gravitational instability within globally stable structures.
% , which are supported by turbulence and rotation on large scale.
On the other hand, we find $\alpha_{\rm vir}\approx$\,2 for the MCs in the satellite galaxies,
which we interpret as result of the weaker stellar feedback and less frequent \SF compared to \flower.
Our results suggest that the more massive and bigger MCs compared to the Milky Way
manifest from the higher gas mass fraction, surface density, and velocity dispersion.
% , which together sets the scale for fragmentation.
High resolution imaging of the first galaxies with the Atacama Large (sub-)Millimeter Array (ALMA) and the
Next Generation Very Large Array (ngVLA)
can furnish scrutinizing data to test our findings and shed light on \SF since the cosmic dark ages.
\end{abstract}
\keywords{methods: data analysis --
          galaxies: high-redshift --
          galaxies: ISM --
          galaxies: evolution --
          galaxies: formation --
          galaxies: starburst --
          stars: formation}

@@

\subsection{Gravitational Instabilities}   \label{sec:Q}
The onset of gravitational instability is considered to be tightly connected to \SF \citep[e.g.,][]{Kennicutt89a, Wang94a, Li05b, Li06a}.
From virial theorem, Jeans 
instability sets the scales at which gravitational potential overcomes kinetic energy. Namely, gravitational instabilities 
dominates at scales $L > L_J$, where $L_J$ is the Jeans length. 
However, in a differentially rotating system, centrifugal force resulting from differential rotation can 
inhibit collapse even on these scales. Thus, a disk is thought to be 
unstable on scales of $L_J < L < L_{\rm rot}$, where $L_{\rm rot}$ is set 
% \propto \Sigma/B^2
by the surface density of the disk and the angular velocity of the rotation.

In theoretical and observational works, gravitational instability has been widely discussed in the 
context of the Toomre stability criterion \citep{Toomre64a, Goldreich65b}, or the Toomre $Q$ parameter 
(or the proxy, circular velocity to dispersion ratio; $v_{\rm circ}/\sigma$; e.g., \citealt{GarciaBurillo03a, Genzel11a, Kassin12a, Leung19a}).
% https://ned.ipac.caltech.edu/level5/March15/Glazebrook/Glazebrook5.html
%
% from linear analysis
For {\it axisymmetric} mode ($m$\eq0), the dispersion relation for the growth of density perturbation in a rotating turbulent disk
of finite thickness $h$ 
is described by:
\begin{equation}
\omega^2 = \kappa^2 - \frac{2\pi G \Sigma |k|}{1 + |k| h} + \sigma^2 k^2,
\label{eqn:3Ddisp}
\end{equation}
where $k$ is the wavenumber and $\kappa$ is the epicyclic frequency, defined as:
\begin{equation}
\kappa^2\equiv\frac{2\Omega}{R}\frac{d}{dR}\left(R^2\Omega\right).
\label{eqn:kappa}
\end{equation}
% \kappa\propto\Omega
The first term on the RHS of Equation~\ref{eqn:3Ddisp} describes rotation support, whereas the second term describes self-gravity, and
the third term describes pressure support.
One can then define the so-called Toomre-Q parameter as:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{\pi G \Sigma}.
\label{eqn:Q}
\end{equation}
For $kh\ll1$, Equation~\ref{eqn:3Ddisp} reduces to the thin disk approximation, which 
becomes unstable % perturbation grows exponentially in time
for some $k$ when $\omega^2 < 0$ (or equivalently, $Q < 1$; \citealt{Toomre64a}).
Under this condition (i.e., when local gravitational instability 
overcomes support from pressure and rotation), fragmentation {\it can} happen. 
% but further evolution depends on their EOS.
%(e.g., instability impose by angular momentum from gas accretion from the cosmic web and
%satellite galaxies and stellar feedback).
Similarly, for a collisionless fluid, the Toomre $Q$ parameter is defined by:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{3.36 G \Sigma}
\end{equation}
\citep{Toomre64a}.

@@

For the purpose of this work, we present Toomre $Q$ map in \Sec{results} and discuss the importance of support against axisymmetric perturbations from galactic rotation and shear in the extracted velocity dispersions of the MCs in \Sec{origin}.
\DL{so does the extracted velocity disp. has a contribution from shear or is it just pressure? I thought we have concluded 
that it is just high pressure leading to the high velocity disp, but not because of diff. rotation. It's not clear to me how 
the Toomre $Q$ map is helpful to inform us on how much of the extracted velo. disp. is coming from diff. rotation.}

@@

The mass of an MC is calculated from the uniformly-gridded 3D density field,
integrating over all zones of the MC, multiplied by its volume.
The size of an MC ($R$) is defined assuming spherical geometry (i.e.,
the size is parameterized via the radius of a sphere, which has a volume corresponding to that of the identified MC).
% {\AP The size is an effective radius, calculated from the total volume of the MC assuming spherical symmetry.}
Since in \obs, dense gas contributes more to the observed linewidths than the more diffuse gas. We therefore calculate
the non-thermal component of the velocity dispersion ($\sigma_{\rm NT}$) of each MC as a density-weighted quantity:
\begin{equation}
\sigma_{\rm NT}^2 = \frac{1}{{3}}\frac{\sum_{i} \rho_i \left|\mathbf{v}_i - \mathbf{\bar v}\right|^2}{\sum_i \rho_i}\,,
\end{equation}
which is summed over the cells composing each MC.
The total velocity dispersion includes contributions from the thermal sound speed ($c_s$):
\begin{equation}
\sigma^2 = \sigma_{\rm NT}^2 + c_s^2.
\end{equation}
The local sound speed of all identified molecular complexes are much smaller compared to the their
turbulent velocities\footnote{We find comparable velocity dispersion between that derived from
taking the root mean square versus that derived from thermal and non-thermal pressure terms, indicating that
rotation velocity is unlikely to be the dominant source of velocity dispersion reported here.
% We further quantify the contributions arising from galactic rotation and shear to the extracted velocity dispersions of the MCs in \Sec{Q}.
In the subsequent sections, we adopt the root mean square of the velocity field
as the velocity dispersion ($\sigma$).}.
On average, the Mach number of the MCs corresponds to about $\mathcal{M}\approx50$.
This is consistent with the global analysis done on \flower by \citet{Vallini18a}.

@@

Maps of the gas Toomre $Q$ parameter and the effective two-component $Q_{\rm eff}$ parameter along different projection angles during the accretion phase of \flower are shown in \Fig{Q} and \Fig{Q_eff}, respectively (see \Fig{h2density} for examples of clumps identified 
in this snapshot).

@@

\subsection{Placing Results in the Context of Existing Observations} \label{sec:diss1}
The population of sub-MCs has dynamics largely
similar to those observed
in $z$\ssim2 spatially resolved studies of gas-rich star-forming galaxies, in
terms of their velocity dispersions, sizes, and gas surface densities (\Fig{larsons_single}; see
e.g., \citealt{Swinbank11a}), with sizes of the order of $R\simeq$\,100\,pc and velocity
dispersions of $\sigma\simeq$\,20$-$80\,\kms.
% velocity dispersion, larsons
Their velocity dispersions are also comparable those observed in the inner Milky way and
nearby gas-rich galaxies (e.g., M64; \citealt{Oka01a, Rosolowsky05a, Heyer09a}), which lies
along the locus of $\sigma\propto R^{0.56}$.
Such high velocity dispersions and surface densities are expected since
they are located in the nuclear regions of the
galaxy, where the potential well is also deeper. The higher velocity dispersions
in these MCs can also be understood as they have experienced more recent episodes
of \SF as \flower is assembling its stellar mass (these MCs also have higher stellar-to-gas mass ratios of $\sim$60).
In fact, by $z\simeq$\,7.2 (i.e., the snapshot corresponding to an early phase of accretion as
shown in \Fig{SFH}a), \flower has assembled
a stellar mass of $M_*$\eq7.5\E{9}\,\Msun. Thus, the higher velocity dispersion
is due in part to the stronger stellar feedback compared to e.g., MCs in its satellite galaxies ($\sigma\approx$10\,\kms).

>>>>

\begin{abstract}
We study the dynamical properties of molecular clouds complexes (MC) in a prototypical galaxy (\flower)
and its satellite galaxies at the Epoch of Reionization (EoR) using cosmological zoom-in simulation.
We identify MCs using H$_2$ density, 
and compare their masses, sizes, velocity dispersions, gas surface densities, and virial parameters ($\alpha_{\rm vir}$)
with those observed in nearby and \z$\sim$\,2 galaxies.
The MCs span a mass range of $10^{6.5-9}$\,\Msun. Their velocity dispersions and gas surface densities are 
systematically higher than those observed in nearby galaxies, but comparable to starburst galaxies.
We compute the effective two-component Toomre $Q$ parameter (gas and stars) to examine 
regions that are unstable to local axisymmetric perturbations. 
We further perform virial analysis, which is accessible from observations, to examine regions likely 
to undergo collapse.
Our analysis indicates that the MCs in the main disk of \flower are unbound, but its 
substructures have lower virial parameters, which is
consistent with the notion that collapsing structures result from gravitational instability within globally stable structures.
% , which are supported by turbulence and rotation on large scale.
On the other hand, we find $\alpha_{\rm vir}\approx$\,2 for the MCs in the satellite galaxies,
which we interpret as result of the weaker stellar feedback and less frequent \SF compared to \flower.
Our results suggest that the more massive and bigger MCs compared to the Milky Way
manifest from the higher gas mass fraction, surface density, and velocity dispersion.
% , which together sets the scale for fragmentation.
High resolution imaging of the first galaxies with the Atacama Large (sub-)Millimeter Array (ALMA) and the
Next Generation Very Large Array (ngVLA)
can furnish scrutinizing data to test our findings and shed light on \SF since the cosmic dark ages.
\end{abstract}
\keywords{methods: data analysis --
          galaxies: high-redshift --
          galaxies: ISM --
          galaxies: evolution --
          galaxies: formation --
          galaxies: starburst --
          stars: formation}

@@

\subsection{Gravitational Instabilities}   \label{sec:Q}
The onset of gravitational instability is considered to be tightly connected to \SF \citep[e.g.,][]{Kennicutt89a, Wang94a, Li05b, Li06a}.
From virial theorem, Jeans 
instability sets the scales at which gravitational potential overcomes kinetic energy. Namely, gravitational instabilities 
dominates at scales $L > L_J$, where $L_J$ is the Jeans length. 
However, centrifugal force resulting from differential rotation in disk galaxies can 
inhibit collapse even on these scales. As a result, a disk is thought to be 
unstable on scales between $L_J < L < L_{\rm rot}$, where $L_{\rm rot}$ is set 
% \propto \Sigma/B^2
by the surface density of the disk and the angular velocity of the rotation.

In theoretical and observational works, gravitational instability has been widely discussed in the 
context of the Toomre stability criterion \citep{Toomre64a, Goldreich65b}, or the Toomre $Q$ parameter 
(or the proxy, circular velocity to dispersion ratio; $v_{\rm circ}/\sigma$; e.g., \citealt{GarciaBurillo03a, Genzel11a, Kassin12a, Leung19a}).
% https://ned.ipac.caltech.edu/level5/March15/Glazebrook/Glazebrook5.html
%
% from linear analysis
For {\it axisymmetric} mode ($m$\eq0), the dispersion relation for the growth of density perturbation in a rotating turbulent disk
of finite thickness $h$ 
is described by:
\begin{equation}
\omega^2 = \kappa^2 - \frac{2\pi G \Sigma |k|}{1 + |k| h} + \sigma^2 k^2,
\label{eqn:3Ddisp}
\end{equation}
where $k$ is the wavenumber and $\kappa$ is the epicyclic frequency, defined as:
\begin{equation}
\kappa^2\equiv\frac{2\Omega}{R}\frac{d}{dR}\left(R^2\Omega\right).
\label{eqn:kappa}
\end{equation}
% \kappa\propto\Omega
The first term on the RHS of Equation~\ref{eqn:3Ddisp} describes rotation support, whereas the second term describes self-gravity, and
the third term describes pressure support.
One can then define the so-called Toomre-Q parameter as:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{\pi G \Sigma}.
\label{eqn:Q}
\end{equation}
For $kh\ll1$, Equation~\ref{eqn:3Ddisp} reduces to the thin disk approximation, which 
becomes unstable % perturbation grows exponentially in time
for some $k$ when $\omega^2 < 0$ (or equivalently, $Q < 1$; \citealt{Toomre64a}).
Under this condition (i.e., when local gravitational instability 
overcomes support from pressure and rotation), fragmentation {\it can} happen. 
% but further evolution depends on their EOS.
%(e.g., instability impose by angular momentum from gas accretion from the cosmic web and
%satellite galaxies and stellar feedback).
% Note that fragmentation is expected to happen at critical scale length 
% $\lambda_{\rm crit} < 2 \pi^2 G \Sigma / \kappa^2$, which is greater than the typical size of 
% GMCs. It should be clear that the instability sets the scales for fragmentation, but 
% star-forming GMCs are the collapsing, denser, and cooler molecular structures on even smaller scales.
% The fragmented structures thus correspond to regions where $Q\lesssim1.
Similarly, for a collisionless fluid, the Toomre $Q$ parameter is defined by:
\begin{equation}
Q\equiv\frac{\sigma\kappa}{3.36 G \Sigma}
\end{equation}
\citep{Toomre64a}.

@@

For the purpose of this work, we present Toomre $Q$ map in \Sec{results} and discuss the 
importance of support from galactic rotation and shear against axisymmetric perturbations in \Sec{origin}.

@@

The mass of an MC is calculated from the uniformly-gridded 3D density field,
integrating over all zones of the MC, multiplied by its volume.
The size of an MC ($R$) is defined assuming spherical geometry (i.e.,
the size is parameterized via the radius of a sphere, which has a volume corresponding to that of the identified MC).
% {\AP The size is an effective radius, calculated from the total volume of the MC assuming spherical symmetry.}
Since in \obs, dense gas contributes more to the observed linewidths than the more diffuse gas. We therefore calculate
the non-thermal component of the velocity dispersion ($\sigma_{\rm NT}$) of each MC as a density-weighted quantity:
\begin{equation}
\sigma_{\rm NT}^2 = \frac{1}{{3}}\frac{\sum_{i} \rho_i \left|\mathbf{v}_i - \mathbf{\bar v}\right|^2}{\sum_i \rho_i}\,,
\end{equation}
which is summed over the cells composing each MC.
The total velocity dispersion includes contributions from the thermal sound speed ($c_s$):
\begin{equation}
\sigma^2 = \sigma_{\rm NT}^2 + c_s^2.
\end{equation}
The local sound speed of all identified molecular complexes are much smaller compared to the their
turbulent velocities\footnote{We find comparable velocity dispersion between that derived from
taking the root mean square versus that derived from thermal and non-thermal pressure terms, indicating that
rotation velocity is unlikely to be the dominant source of velocity dispersion reported here.
% We further quantify the contributions arising from galactic rotation and shear to the extracted velocity dispersions of the MCs in \Sec{Q}.
In the subsequent sections, we adopt the root mean square of the velocity field
as the velocity dispersion ($\sigma$). \label{ftn:veldisp}}.
On average, the Mach number of the MCs corresponds to about $\mathcal{M}\approx50$.
This is consistent with the global analysis done on \flower by \citet{Vallini18a}.

@@

Maps of the gas Toomre $Q$ parameter and the effective two-component $Q_{\rm eff}$ parameter along different projection angles during the accretion phase of \flower are shown in \Fig{Q} and \Fig{Q_eff}, respectively (see \Fig{h2density} for examples of clumps identified 
in this snapshot). Fragmentation are expected in regions of low-$Q$. 
Further interpretation and implication of this is discussed in \Sec{origin}.

@@

\subsection{Placing Results in the Context of Existing Observations} \label{sec:diss1}
The population of sub-MCs has dynamics largely
similar to those observed
in $z$\ssim2 spatially resolved studies of gas-rich star-forming galaxies, in
terms of their velocity dispersions, sizes, and gas surface densities (\Fig{larsons_single}; see
e.g., \citealt{Swinbank11a}), with sizes of the order of $R\simeq$\,100\,pc and velocity
dispersions of $\sigma\simeq$\,20$-$80\,\kms.
% velocity dispersion, larsons
Their velocity dispersions are also comparable to those observed in the inner Milky way and
nearby gas-rich galaxies (e.g., M64; \citealt{Oka01a, Rosolowsky05a, Heyer09a}), which lies
along the locus of $\sigma\propto R^{0.56}$.
Such high velocity dispersions and surface densities are expected since
they are located in the nuclear regions of the
galaxy, where the potential well is also deeper. The higher velocity dispersions (see also 
footnote~\ref{ftn:veldisp} on page \pageref{ftn:veldisp})
of these MCs can also be understood as they have experienced more recent episodes
of \SF as \flower is assembling its stellar mass (these MCs also have higher stellar-to-gas mass ratios of $\sim$60).
In fact, by $z\simeq$\,7.2 (i.e., the snapshot corresponding to an early phase of accretion as
shown in \Fig{SFH}a), \flower has assembled
a stellar mass of $M_*$\eq7.5\E{9}\,\Msun. Thus, the higher velocity dispersion
is due in part to the stronger stellar feedback compared to e.g., MCs in its satellite galaxies ($\sigma\approx$10\,\kms).
